---
title: "Tornadoes"
output: html_document
date: "`r Sys.Date()`"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(correlation)
library(maps)
library(sjPlot)

theme_set(theme_light())
```

# Read data

We read the dataset from the github.

```{r}
tornados_raw <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2023/2023-05-16/tornados.csv')

```

# Process data

```{r}
tornados <- 
    tornados_raw |> 
    mutate(mo = factor(mo, labels = month.name),
           lg_loss = log(loss + 1))


```

Plot tornado frequency by date

```{r}
tornados |> 
    count(yr) |> 
    ggplot() +
    aes(x = yr, y = n) +
    geom_col() +
    geom_smooth(method = "lm")
    
tornados |> 
    mutate(mo = factor(mo, labels = month.name)) |> 
    ggplot() +
    aes(x = mo) +
    geom_bar()

```

# Correlation of monetary loss, insjuries, fatalities, and magnitude

```{r}
    
tornados |> 
    select(mag:loss) |> 
    correlation()

tornados |> 
    select(mag:loss) |> 
    correlation() |> 
    summary()

tornados |> 
    select(mag:loss) |> 
    correlation() |> 
    plot()

```


# Plot tornado frequency by state

```{r}
tornados |> 
    count(st, sort = TRUE)

states <- tibble(st = state.abb,
                 region = str_to_lower(state.name))

tornados |> 
    count(st) |> 
    left_join(states, by = "st") |> 
    left_join(map_data("state"), by = "region") |> 
    group_by(st) |> 
    mutate(avg_long = mean(long),
           avg_lat = mean(lat)) |> 
    ggplot() +
    aes(x = long, y = lat, group = group, fill = n) |> 
    geom_polygon(color = "white") +
    geom_text(aes(x = avg_long, y = avg_lat, label = st), color = "grey") +
    coord_quickmap() + 
    scale_fill_viridis_c() +
    theme_void() +
    labs(title = "Frequency of tornadoes by state")


```

# Plot tornado origin and destination on map (failed)

```{r}
qplot(tornados$mag)

tornados |> 
    count(mag)



tornados |> 
    count(st) |> 
    left_join(states, by = "st") |> 
    left_join(map_data("state"), by = "region") |> 
    group_by(st) |> 
    mutate(avg_long = mean(long),
           avg_lat = mean(lat)) |> 
    ggplot() +
    aes(x = long, y = lat, group = group, fill = n) |> 
    geom_polygon(color = "white") +
    geom_text(aes(x = avg_long, y = avg_lat, label = st), color = "grey") +
    coord_quickmap() + 
    scale_fill_viridis_c() +
    theme_void() +
    labs(title = "Frequency of tornadoes by state")
```


```{r}
big_tornadoes <- 
    tornados |> 
    filter(elat != 0, mag == 5) |> 
    mutate(size = len * wid)

tornados |> 
    count(st) |> 
    left_join(states, by = "st") |> 
    left_join(map_data("state"), by = "region") |> 
    group_by(st) |> 
    mutate(avg_long = mean(long),
           avg_lat = mean(lat)) |> 
    ggplot() +
    aes(x = long, y = lat, group = group) |> 
    geom_polygon(color = "white", fill = "lightgray") +
    geom_text(aes(x = avg_long, y = avg_lat, label = st), color = "darkgray") +
    geom_segment(data = big_tornadoes, 
                 aes(x = slon, y = slat, xend = elon, yend = elat, linewidth = size),
                 arrow = arrow(angle = 15, length = unit(0.2, "cm"), type = "closed"), 
                 alpha = .5, color = "blue") +
    coord_quickmap() + 
    theme_void() +
    scale_linewidth_continuous(labels = scales::comma_format()) +
    labs(title = "Trajectories of the biggest tornadoes",
         linewidth = "Size of the tornado (m2)")



```

# Stat modeling of monetary loss

Building two models, one with categorical predictors (state, month), and one without.

```{r}
tornado_lm <- lm(lg_loss ~ yr + mo + st + mag + len + wid, data = tornados)
tornado_ns_lm <- lm(lg_loss ~ yr + mag + len + wid, data = tornados)
```


```{r}
summary(tornado_lm)

anova(tornado_lm)

tab_model(tornado_lm, 
          show.std = TRUE,
          show.aic = TRUE, show.fstat = TRUE, show.reflvl = TRUE)

parameters::parameters(tornado_lm) |> 
    print_html()
```


## Investigating assumotions

```{r}
performance::check_model(tornado_ns_lm)
```

## Comparing models
```{r}
anova(tornado_lm, tornado_ns_lm)

tab_model(tornado_lm, tornado_ns_lm,
          show.aic = TRUE)

```

